Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-07-20 MIMOSA.tex
Updated fluidigm plots
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\section{Introduction}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize and subset these cells. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references). This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
\node[anchor=south west, inner sep=0] at (4.6,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot3.pdf}};
\node at (0,3) [font=\tiny\sffamily] {A} ;
\node at (4.5,3) [font=\tiny\sffamily] {B} ;
\node at (8.95,3) [font=\tiny\sffamily] {C} ;
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a significant decrease, green a significant increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The $log_2$ ratio of the proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. C) The difference in the proportion of cells expressing each gene in the stimulated vs. control samples. Ordering of the rows and columns is preserved as in A). The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).}

\label{fig:fluidigm}
\end{figure}

@@

HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}. The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}. Vaccine was given at two increasing doses, as well as a placebo. T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}. The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}. The sample size consisted of 20 vaccine and four placebo recipients. The original statistical analysis of the positivity calls is described in the associated publication~\cite{Peiperl:2010ej}. The Gag stimulated, IL2 expressing, CD4+ T--cell data from day 28 was used to derive hyper--parameter estimates for the simulation studies.

>>>>

\section{Introduction}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references). This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
\node[anchor=south west, inner sep=0] at (4.6,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/FluidigmFisher.pdf}};
\node at (0,3) [font=\tiny\sffamily] {A} ;
\node at (4.5,3) [font=\tiny\sffamily] {B} ;
\node at (8.95,3) [font=\tiny\sffamily] {C} ;
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The posterior differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Genes selected from Fisher's exact test at the 10\% FDR level. Clustering of the maximum likelihood estimates of difference in proportions of cells expressing a gene in stimulated vs. control samples with significant genes selected using Fisher's exact test.}
\label{fig:fluidigm}
\end{figure}

@@

%HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}. The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}. Vaccine was given at two increasing doses, as well as a placebo. T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}. The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}. The sample size consisted of 20 vaccine and four placebo recipients. The original statistical analysis of the positivity calls is described in the associated publication~\cite{Peiperl:2010ej}. The Gag stimulated, IL2 expressing, CD4+ T--cell data from day 28 was used to derive hyper--parameter estimates for the simulation studies.
